---
title: "Multilevel Modelling"
subtitle: Lecture 5 (R code)
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### Preliminaries

This auxiliary notebook reproduces the analyses carried out in the first part of Lecture 5 (excluding the 3-level part). 

We load the R packages *haven*,  *ggplot2*, *lme4*, and *lmerTest*. 

```{r}
require(haven)      # to load the SPSS .sav file
require(ggplot2)    # used for production of graphs
require(lme4)       # used for fitting random effect models
require(lmerTest)   # required for ranova
```

We load and prepare the data as previously:

```{r}
pop.rawdata <- 
  read_sav(file="https://github.com/MultiLevelAnalysis/Datasets-third-edition-Multilevel-book/blob/master/chapter%202/popularity/SPSS/popular2.sav?raw=true")
``` 

```{r}
pop.data <- pop.rawdata[,c("pupil", "class", "extrav", "sex", "texp", "popular")]

colnames(pop.data)<- c("pupil", "class", "extraversion", "gender", "experience", "popularity")

head(pop.data) # we have a look at the first 6 observations
```

### Models with and without random slope 

```{r}
model0 <- lmer(formula = popularity ~  1+ extraversion +(1|class), data    = pop.data) 
```

```{r}
model1 <- 
  lmer(
    formula = popularity ~  1+extraversion +( extraversion|class), data = pop.data) 
summary(model1)
```

### Deviance (LR) test

```{r}
 deviance(model0)    # works but is deprecated
-2*summary(model0)$logLik
```

```{r}
# deviance(model1)
-2*summary(model1)$logLik
```

```{r}
# deviance(model0)-deviance(model1)
-2*summary(model0)$logLik + 2*summary(model1)$logLik 
qchisq(0.95,2)
```


```{r}
ranova(model1)
-2889.7*2
-2916.3*2
```

### More than one predictor variable

```{r}
model2 <- 
  lmer(formula = popularity ~ 1 + gender + extraversion + experience + 
         (1 + gender + extraversion | class), data = pop.data)
summary(model2)
```

```{r}
model3 <- 
  lmer(formula = popularity ~ 1 + gender + extraversion + experience + 
         (1 + extraversion | class), data = pop.data)

deviance(model3)-deviance(model2)

qchisq(0.95,3)
```

### Fitted model summaries

```{r}
summary(model3)
names(summary(model3))
```

```{r}
summary(model3)$varcor
```

```{r}
summary(model3)$vcov
round(summary(model3)$coef, digits=4)
round(sqrt(diag(summary(model3)$vcov)), digits=4)
```

### Fitted models by gender

```{r}
pop.data$pred3 <- predict(model3)
gender_names <- c(`0`="Male",`1`="Female") #define labels for facet_grid

ggplot(pop.data,
       aes(x = extraversion, y = popularity, col = class, group = class)
       ) +
       geom_line(aes(y=pred3,group=class, col=class)) +
       scale_color_gradientn(colours = rainbow(100)) +
        facet_grid(.~gender, labeller = as_labeller(gender_names))
```

### Bonus material: Diagnostics

```{r}
plot(model3)
```

That’s good --- no pattern at all. The model is correctly specified (in terms of the included predictor variables).


```{r}
qqnorm(resid(model3))
qqline(resid(model3), col = "red") 
```

Almost perfectly follows straight line.
Can conclude that residual distribution is normal, confirming that the assumption of normality of the errors is reasonable.

```{r}
qqnorm(ranef(model3)$class[,1] )
qqline(ranef(model3)$class[,1], col = "red")
```

About ok...

```{r}
qqnorm(ranef(model3)$class[,2])
qqline(ranef(model3)$class[,2], col = "red")
```

Ok.

