---
title: "Multilevel Modeling Summative"
author: "Anonymous Marking Code: Z0195806"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    latex_engine: xelatex
    toc: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1 Introduction

## 1.1 Background of Multisite Trials

**Definition of Multisite Trials**: Multisite trials are a type of clinical research study where the intervention being tested is administered across multiple sites or locations. These trials are particularly valuable in assessing the effectiveness of an intervention in a broader, more diverse population. By including a variety of settings, such as different hospitals, clinics, or communities, multisite trials can provide more generalizable results, ensuring that the findings are not specific to a single location or population [@YouthEndowmentFund2024].

## 1.2 Intro to the MST Dataset

```{r warning=FALSE,message=FALSE}
# --------------------------------------------------------
## clear the environment var area
# rm(list = ls())
## clear all plots
# graphics.off()
## clear the console area
# cat("\014")
# --------------------------------------------------------
# install.packages("gridExtra")
# --------------------------------------------------------
require(lme4)
require(lmerTest)
require(ggplot2)
require(sjPlot)
```

Download the dataset "MST" only once from GitHub and save it to csv files.

```{r}
# MST <-
#   read.csv(
#     "https://andygolightly.github.io/teaching/MATH43515/summative/andy.csv",
#     header = TRUE
#   )
# write.csv(MST, "MST.csv")
MST = read.csv("./MST.csv")
head(MST)
# dim(MST)
## Show three line table MST with sjPlot::tab_df
# tab_df(MST[1:5, ])
```

```{r}
library(ggplot2)
```

```{r message=FALSE}
# Correlation between two varibales with GGpairs
library("GGally")
ggpairs(MST)+theme_bw()
```

# Part 2 Methods

# Part 3 Analysis

# Part 4 Discussion of results

# Word Count

```{r message=FALSE, warning=FALSE}
# install.packages("devtools")
# devtools::install_github("benmarwick/wordcountaddin",
#                          type = "source", dependencies = TRUE)
require(wordcountaddin)
word_count()
text_stats()
```

# References
